---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
library(DESeq2)
library(ggplot2)
library(vsn)
library(pheatmap)
library(ggrepel)
library(ggpubr)

source("https://raw.githubusercontent.com/NUPulmonary/utils/master/R/plotPCA_manual.R")

source("https://github.com/NUPulmonary/utils/raw/master/R/go_enrichment.R")
```

Set random seed
```{r}
set.seed(1066)
```


```{r}
plot_lib_sizes <- function(data) {
    df <- data.frame(lib.size = sort(colSums(data)))
        ggplot(df, aes(x = reorder(rownames(df), lib.size), y = lib.size)) +
        geom_col() +
        ggtitle("Sorted library sizes") +
        theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1)) +
        xlab("")
}
```


# Nasal samples from CF

We have samples from 3 genotypes: Control, F508del homozygous and F508del heterozygous.
We name them: control, F508del and CompHet respectively (compound heterozygotes).

Also we have 2 samples from universal human RNA as control samples.

Todo:
1. PCA analysis
2. DEGs between 3 groups with k-means clustering


### Load and preprocess data

```{r}
data <- read.table("Validation_cohort_raw_counts_relabeled_for_upload_with_genes.txt", header = TRUE)
```

Pick only samples
```{r}
coldata <- data.frame(row.names = colnames(data)[3:ncol(data)])
```

```{r}
rownames(coldata) <- sub("^X", "", rownames(coldata))
```

Prepare sample metadata:
sample_type = sample / uhRNA
genotype = uhRNA / control / F508del / CompHet

```{r}
coldata$sample_type <- "sample"
coldata$sample_type[grepl("uhrna", rownames(coldata))] <- "uhRNA"
```

```{r}
coldata$genotype <- "none"
coldata$genotype[grepl("uhrna", rownames(coldata))] <- "uhRNA"
coldata$genotype[grepl("Control", rownames(coldata))] <- "HV"
coldata$genotype[grepl("F508del", rownames(coldata))] <- "F508del"
coldata$genotype[grepl("CompHet", rownames(coldata))] <- "CompHet"
```

Separate counts
```{r}
counts <- as.matrix(data[, 3:ncol(data)])
colnames(counts) <- sub("^X", "", colnames(counts))
rownames(counts) <- data$Ensembl_ID
```

Now look! In the whole table there are no odd numbers! Isn't it strange…
```{r}
sum(counts %% 2 == 1)
```

```{r}
plot_lib_sizes(counts / 2)
```


I'll just divide all counts by 2 then
```{r}
dds <- DESeqDataSetFromMatrix(counts / 2, coldata, ~ genotype)
```

Add gene info
```{r}
rowData(dds) <- data[, 1:2]
```

Remove genes with less than 11 counts across all samples
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
```

### Check quality and structure

Compute distances between samples
```{r}
sampleDists <- dist(t(assay(vst(dds))))
```

```{r}
pheatmap(as.matrix(sampleDists), cluster_cols = FALSE, cluster_rows = FALSE, show_colnames = FALSE)
```

uhRNA cluster together and different from everything else, good!
Some structure visible in Control group, good.
CompHet and F508del groups seem to be split between themselves.

Sample 4 seems to be an outlier.


Let's plot PCA
```{r}
d <- plotPCA(vst(dds), intgroup = c("genotype"), returnData = TRUE)
percentVar <- round(100 * attr(d, "percentVar"))

ggplot(d, aes(x = PC1, y = PC2, color = genotype)) +
  geom_point() + 
  geom_label_repel(aes(label = sapply(colnames(dds), substr, 1, 2)), show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed()
```
uhRNA are together and a bit apart from everything else, good!. Sample 4 is on the left.

Let's remove it and see what happens.


```{r}
counts2 <- counts[, -4]
coldata2 <- coldata[-4, ]
```

```{r}
dds <- DESeqDataSetFromMatrix(counts2 / 2, coldata2, ~ genotype)
```

```{r}
rowData(dds) <- data[, 1:2]
```

```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
```

```{r}
sampleDists <- dist(t(assay(vst(dds))))
```

```{r}
pheatmap(as.matrix(sampleDists), cluster_cols = FALSE, cluster_rows = FALSE, show_colnames = FALSE)
```
Seems better.

Check PCA

```{r}
d <- plotPCA(vst(dds), intgroup = c("genotype"), returnData = TRUE)
percentVar <- round(100 * attr(d, "percentVar"))

ggplot(d, aes(x = PC1, y = PC2, color = genotype)) +
  geom_point() + 
  geom_label_repel(aes(label = sapply(colnames(dds), substr, 1, 2)), show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed()
```
A little bit better.

### Remove uhRNA samples and move to DEGs

```{r}
ncol(counts2)
```


```{r}
counts3 <- counts2[, c(-35, -36)]
coldata3 <- coldata2[c(-35, -36), ]
```

```{r}
rownames(counts3) <- data$Ensembl_ID
```


```{r}
dds <- DESeqDataSetFromMatrix(counts3 / 2, coldata3, ~ genotype)
```

```{r}
rowData(dds) <- data[, 1:2]
```

```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]
```

```{r}
sampleDists <- dist(t(assay(vst(dds))))
```

```{r}
pheatmap(as.matrix(sampleDists), cluster_cols = FALSE, cluster_rows = FALSE, show_colnames = FALSE)
```

Let's see normalization methods

Log-norm
```{r}
meanSdPlot(assay(normTransform(dds)))
```

VST with parametric dispersions
```{r}
meanSdPlot(assay(vst(dds, blind = FALSE)))
```

VST with local dispersions
```{r}
meanSdPlot(assay(vst(dds, blind = FALSE, fitType = "local")))
```

Let's use local dispersion estimations

PCA
```{r}
d <- plotPCA(
  vst(dds, blind = FALSE, fitType = "local"), 
  ntop = 2000,
  intgroup = c("genotype"), 
  returnData = TRUE
)
percentVar <- round(100 * attr(d, "percentVar"))

ggplot(d, aes(x = PC1, y = PC2, color = genotype)) +
  geom_point() + 
  geom_label_repel(aes(label = sapply(colnames(dds), function(x) {strsplit(x, "_")[[1]][1]})), show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed()
```
No clear structure. PC1 is most of variation, PC2 is much lower.
Controls seem to be in lower right corner, while F508del and CompHet are spread out.

Save as PDF
```{r}
pdf("01pca.pdf", width=8, height=6)
ggplot(d, aes(x = PC1, y = PC2, color = genotype)) +
  geom_point() + 
  geom_label_repel(aes(label = sapply(colnames(dds), function(x) {strsplit(x, "_")[[1]][1]})), show.legend = FALSE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  coord_fixed()
dev.off()
```


Let's look at loadings of first PCs
```{r}
d <- plotPCA_manual(
  vst(dds, blind = FALSE, fitType = "local"), 
  intgroup = c("genotype"), 
  ntop = 2000,
  pcs = 10
)
percentVar <- round(100 * attr(d$data, "percentVar"))
```

```{r}
loadings <- data.frame(d$pca$rotation)
genes <- rownames(loadings)
gene_table <- rowData(dds)
for (i in 1:7) {
  print(sprintf("PC%d: %d %% variance", i, percentVar[i]))
  l <- loadings[, i]
  ord <- order(abs(l), decreasing = TRUE)
  g <- genes[ord][1:20]
  v <- paste(gene_table[g, "Gene_Name"], " (", round(l[ord][1:20], 4), ")")
  print(paste(v, collapse = ", "))
  print("=============")
}
```

Well, loadings seem to be:
PC1: monocytes
PC2: ciliated cells
PC3: ????
PC4: epithelial repair (???)
PC5: inflammation


Let's do K-means!

First, LRT between 3 conditions
```{r}
dds <- DESeq(dds, fitType = "local", test = "LRT", reduced = ~ 1)
```

```{r}
deseq_results <- as.data.frame(results(dds))
```

Now select significant genes
```{r}
genes_of_interest <- rownames(subset(deseq_results, padj < 0.05))
```

```{r}
counts_mat_orig <- counts(dds, normalized = TRUE)
dds_orig <- dds
counts_mat <- counts_mat_orig[genes_of_interest, ]
```

```{r}
counts_mat <- t(scale(t(counts_mat)))
```

I tried 6, 4 and 3
```{r}
k <- 3
```

```{r}
kmeans_results <- as.data.frame(kmeans(x = counts_mat,
                          centers = k, 
                          iter.max = 1000, 
                          nstart = 500)$cluster)
```

```{r}
colnames(kmeans_results) <- "cluster"
```

```{r}
kmeans_results <- kmeans_results[order(kmeans_results$cluster), , drop = FALSE]
counts_mat <- counts_mat[rownames(kmeans_results), ]
```

```{r}
gaps <- cumsum(table(kmeans_results))[1:k]
```

```{r}
anno_col <- data.frame(colData(dds))[, c("genotype"), drop = FALSE]
```

```{r}
plot <- pheatmap(counts_mat, 
                  cluster_rows = FALSE,
                  cluster_cols = TRUE,
                  clustering_method = "ward.D2",
                  gaps_row = gaps,
                  show_colnames = TRUE,
                  show_rownames = FALSE,
                  labels_col = sapply(colnames(dds), function(x) {strsplit(x, "_")[[1]][1]}),
                  annotation_col = anno_col,
                  cutree_cols = 3,
                  border_color = NA,
                  angle_col = 90,
                  annotation_colors = list("genotype" = c(
                    "HV" = "blue",
                    "F508del" = "darkred",
                    "CompHet" = "darkorange3"
                  )))
```

Save as PDF
```{r}
pdf("02kmeans.pdf", width=8, height=6)
print(plot)
dev.off()
```

```{r}
write.csv(plot$tree_col$labels[plot$tree_col$order], "02kmeans-sample-order.csv")
```


```{r}
kmeans_genes <- deseq_results[rownames(kmeans_results),]
kmeans_genes$cluster <- kmeans_results[rownames(kmeans_genes), "cluster"]
```

Print top genes per cluster (lowest p-value)

```{r}
for (i in 1:k) {
  print(sprintf("K-means cluster %d (%d)", i, sum(kmeans_genes$cluster == i)))
  print("Lowest p-value genes")
  this <- kmeans_genes[kmeans_genes$cluster == i,]
  l <- this$padj
  ord <- order(l, decreasing = FALSE)
  g <- rownames(this)[ord][1:20]
  v <- paste(gene_table[g, "Gene_Name"], " (", round(l[ord][1:20], 4), ")")
  print(paste(v, collapse = ", "))
  print("=============")
}
```

Save the table
```{r}
kmeans_genes$gene_name <- gene_table[rownames(kmeans_genes), "Gene_Name"]
write.csv(kmeans_genes, "02kmeans-genes.csv")
```

Cut kmeans sample groups
```{r}
kmeans.groups <- cutree(plot$tree_col, k = 3)
```

```{r}
table(kmeans.groups)
```

```{r}
group.names <- c("1" = "B", "2" = "A", "3" = "C")
```

```{r}
kmeans.samples <- group.names[kmeans.groups]
names(kmeans.samples) <- names(kmeans.groups)
write.csv(kmeans.samples, "02kmeans-sample-groups.csv")
```


Run GO enrichment for each cluster
```{r}
genes_cluster1 <- rownames(kmeans_genes)[kmeans_genes$cluster == 1]
go_cluster1 <- go_enrichment(dds_orig, genes_cluster1, go_annotations = "org.Hs.eg.db")
```

```{r}
genes_cluster2 <- rownames(kmeans_genes)[kmeans_genes$cluster == 2]
go_cluster2 <- go_enrichment(dds_orig, genes_cluster2, go_annotations = "org.Hs.eg.db")
```

```{r}
genes_cluster3 <- rownames(kmeans_genes)[kmeans_genes$cluster == 3]
go_cluster3 <- go_enrichment(dds_orig, genes_cluster3, go_annotations = "org.Hs.eg.db")
```

```{r}
go_cluster1$cluster <- 1
go_cluster2$cluster <- 2
go_cluster3$cluster <- 3
```

```{r}
go_processes <- rbind(go_cluster1, go_cluster2, go_cluster3)
```

```{r}
write.csv(go_processes, "02kmeans-go-processes.csv")
```


## Pairwise comparisons

k-means and deconvolution both identified three groups of samples: “epithelium-rich” cluster, containing controls, CF and heterozygotes, and two immune clusters, mostly containing CF and hets. 

Let's run pairwise DEGs between 1) controls vs CF, 2) control vs hets, 3) CF vs hets using all samples. 

```{r}
dds <- DESeq(dds, fitType = "local")
```
### Control vs CF
```{r}
ctrl.vs.cf <- as.data.frame(results(dds, contrast = c("genotype", "HV", "F508del")))
ctrl.vs.cf <- ctrl.vs.cf[!is.na(ctrl.vs.cf$padj), ]
ctrl.vs.cf$sign <- ""
ctrl.vs.cf$sign[(ctrl.vs.cf$log2FoldChange < 0) & (ctrl.vs.cf$padj < 0.05)] <- "Downregulated"
ctrl.vs.cf$sign[(ctrl.vs.cf$log2FoldChange > 0) & (ctrl.vs.cf$padj < 0.05)] <- "Upregulated"
ctrl.vs.cf <- ctrl.vs.cf[order(ctrl.vs.cf$sign),]
```

```{r}
ace2_id <- gene_table$Ensembl_ID[gene_table$Gene_Name %in% c("ACE2", "TMPRSS2")]
```

```{r}
ace2_data <- as.data.frame(merge(ctrl.vs.cf[ace2_id, ], gene_table[ace2_id, ], by = "row.names"))
```

```{r}
ggplot(ctrl.vs.cf, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
  geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 5000,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) +
  scale_fill_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 6e+04, y = 5, label = sprintf("Up in HV: %d", sum(ctrl.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 6e+04, y = -6, label = sprintf("Up in homozygous\nF508del: %d", sum(ctrl.vs.cf$sign == "Downregulated"))) + 
  ggtitle("HV vs CF F508del") +
  theme(legend.position = "none")
```

Save as PDF
```{r}
pdf("03control-vs-f508del.pdf", width=6, height = 6)
ggplot(ctrl.vs.cf, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
  geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 5000,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) +
  scale_fill_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 6e+04, y = 5, label = sprintf("Up in HV: %d", sum(ctrl.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 6e+04, y = -6, label = sprintf("Up in homozygous\nF508del: %d", sum(ctrl.vs.cf$sign == "Downregulated"))) + 
  ggtitle("HV vs CF F508del") +
  theme(legend.position = "none")
dev.off()
```



Let's verify the sign. Select most expressed upregulated gene
```{r}
upreg <- ctrl.vs.cf[ctrl.vs.cf$log2FoldChange > 0,]
upreg[order(upreg$baseMean, decreasing = TRUE), ][1:5, ]
```

```{r}
expr.df <- coldata3
expr.df$MUC5AC <- counts_mat_orig[rownames(gene_table)[gene_table$Gene_Name == "MUC5AC"], ]
```

And plot it by genotype
```{r}
ggplot(expr.df) + geom_boxplot(aes(x = genotype, y = MUC5AC)) +
  geom_jitter(aes(x = genotype, y = MUC5AC), height = 0, width = 0.1)
```

The sign is correct. Let's save results
```{r}
sign.only <- ctrl.vs.cf[ctrl.vs.cf$padj < 0.05, ]
sign.only$gene_name <- gene_table[rownames(sign.only), "Gene_Name"]
write.csv(sign.only[order(sign.only$log2FoldChange, decreasing = TRUE), ], "03ctrl-vs-cf.csv")
```

Here in control we see T cell signature (CD8A, CD3E etc), but it's not picked up by deconvolution.


### Illustrate k-means cluster composition bias
```{r}
ciliated_secr_markers <- c("CAPS", "TPPP3", "BPIFA1")
cil.df <- expr.df
cil.df$MUC5AC <- NULL
cil.df$cluster <- group.names[kmeans.groups]
cil.df$cluster <- factor(cil.df$cluster, levels=c("A", "B", "C"))

for (g in ciliated_secr_markers) {
  cil.df[[g]] <- counts_mat_orig[rownames(gene_table)[gene_table$Gene_Name == g], ]
}

cil.df.long <- cil.df %>%
  tidyr::pivot_longer(ciliated_secr_markers, names_to = "gene")

ggplot(cil.df.long) + geom_boxplot(aes(x = cluster, y = value), outlier.shape = NA) +
  geom_jitter(aes(x = cluster, y = value, shape = cluster), height = 0, width = 0.2, size = 2) +
  facet_wrap(vars(gene), scales = "free_y") +
  ylab("Normalized expression")
```

Save as PDF
```{r}
pdf("02kmeans-ciliated.pdf", width = 6, height = 3)
ciliated_secr_markers <- c("CAPS", "TPPP3", "BPIFA1")
cil.df <- expr.df
cil.df$MUC5AC <- NULL
cil.df$cluster <- group.names[kmeans.groups]
cil.df$cluster <- factor(cil.df$cluster, levels=c("A", "B", "C"))

for (g in ciliated_secr_markers) {
  cil.df[[g]] <- counts_mat_orig[rownames(gene_table)[gene_table$Gene_Name == g], ]
}

cil.df.long <- cil.df %>%
  tidyr::pivot_longer(ciliated_secr_markers, names_to = "gene")

ggplot(cil.df.long) + geom_boxplot(aes(x = cluster, y = value), outlier.shape = NA) +
  geom_jitter(aes(x = cluster, y = value, shape = cluster), height = 0, width = 0.2, size = 2) +
  facet_wrap(vars(gene), scales = "free_y") +
  ylab("Normalized expression")
dev.off()
```



```{r}
immune_markers <- c("CSF3R", "FCGR3B", "ITGA2")
imm.df <- expr.df
imm.df$MUC5AC <- NULL
imm.df$cluster <- group.names[kmeans.groups]
imm.df$cluster <- factor(imm.df$cluster, levels=c("A", "B", "C"))

for (g in immune_markers) {
  imm.df[[g]] <- counts_mat_orig[rownames(gene_table)[gene_table$Gene_Name == g], ]
}

imm.df.long <- imm.df %>%
  tidyr::pivot_longer(immune_markers, names_to = "gene")

ggplot(imm.df.long) + geom_boxplot(aes(x = cluster, y = value), outlier.shape = NA) +
  geom_jitter(aes(x = cluster, y = value, shape = cluster), height = 0, width = 0.1, size = 2) +
  facet_wrap(vars(gene), scales = "free_y") +
  ylab("Normalized expression")
```

Save as PDF
```{r}
pdf("02kmeans-immune.pdf", width = 6, height = 3)
immune_markers <- c("CSF3R", "FCGR3B", "ITGA2")
imm.df <- expr.df
imm.df$MUC5AC <- NULL
imm.df$cluster <- group.names[kmeans.groups]
imm.df$cluster <- factor(imm.df$cluster, levels=c("A", "B", "C"))

for (g in immune_markers) {
  imm.df[[g]] <- counts_mat_orig[rownames(gene_table)[gene_table$Gene_Name == g], ]
}

imm.df.long <- imm.df %>%
  tidyr::pivot_longer(immune_markers, names_to = "gene")

ggplot(imm.df.long) + geom_boxplot(aes(x = cluster, y = value), outlier.shape = NA) +
  geom_jitter(aes(x = cluster, y = value, shape = cluster), height = 0, width = 0.1, size = 2) +
  facet_wrap(vars(gene), scales = "free_y") +
  ylab("Normalized expression")
dev.off()
```




### Control vs CompHet
```{r}
ctrl.vs.het <- as.data.frame(results(dds, contrast = c("genotype", "HV", "CompHet")))
ctrl.vs.het <- ctrl.vs.het[!is.na(ctrl.vs.het$padj), ]
ctrl.vs.het$sign <- ""
ctrl.vs.het$sign[(ctrl.vs.het$log2FoldChange < 0) & (ctrl.vs.het$padj < 0.05)] <- "Downregulated"
ctrl.vs.het$sign[(ctrl.vs.het$log2FoldChange > 0) & (ctrl.vs.het$padj < 0.05)] <- "Upregulated"
ctrl.vs.het <- ctrl.vs.het[order(ctrl.vs.het$sign),]
```

```{r}
ace2_data <- as.data.frame(merge(ctrl.vs.het[ace2_id, ], gene_table[ace2_id, ], by = "row.names"))
```

```{r}
ggplot(ctrl.vs.het, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
  geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 5000,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  scale_fill_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 6e+04, y = 4, label = sprintf("Up in HV: %d", sum(ctrl.vs.het$sign == "Upregulated"))) +
  annotate("text", x = 6e+04, y = -6, label = sprintf("Up in CompHet: %d", sum(ctrl.vs.het$sign == "Downregulated"))) + 
  ggtitle("HV vs CompHet") +
  theme(legend.position = "none")
```

Save as PDF
```{r}
pdf("03control-vs-comphet.pdf", width=6, height=6)
ggplot(ctrl.vs.het, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
  geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 5000,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  scale_fill_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 6e+04, y = 4, label = sprintf("Up in HV: %d", sum(ctrl.vs.het$sign == "Upregulated"))) +
  annotate("text", x = 6e+04, y = -6, label = sprintf("Up in CompHet: %d", sum(ctrl.vs.het$sign == "Downregulated"))) + 
  ggtitle("HV vs CompHet") +
  theme(legend.position = "none")
dev.off()
```


```{r}
sign.only <- ctrl.vs.het[ctrl.vs.het$padj < 0.05, ]
sign.only$gene_name <- gene_table[rownames(sign.only), "Gene_Name"]
write.csv(sign.only[order(sign.only$log2FoldChange, decreasing = TRUE), ], "03ctrl-vs-het.csv")
```


### CompHet vs CF
```{r}
het.vs.cf <- as.data.frame(results(dds, contrast = c("genotype", "CompHet", "F508del")))
het.vs.cf <- het.vs.cf[!is.na(het.vs.cf$padj), ]
het.vs.cf$sign <- ""
het.vs.cf$sign[(het.vs.cf$log2FoldChange < 0) & (het.vs.cf$padj < 0.05)] <- "Downregulated"
het.vs.cf$sign[(het.vs.cf$log2FoldChange > 0) & (het.vs.cf$padj < 0.05)] <- "Upregulated"
het.vs.cf <- het.vs.cf[order(het.vs.cf$sign),]
```

```{r}
ggplot(het.vs.cf) + 
  geom_point(aes(x = baseMean, y = log2FoldChange, col = sign)) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "red", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 5e+04, y = 5, label = sprintf("Up in CompHet: %d", sum(het.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 5e+04, y = -6, label = sprintf("Down in CompHet: %d", sum(het.vs.cf$sign == "Downregulated"))) + 
  ggtitle("CompHet vs CF F508del")
```
No DEGs here!

Save as PDF
```{r}
pdf("03comphet-vs-f508del.pdf", width=8, height=6)
ggplot(het.vs.cf) + 
  geom_point(aes(x = baseMean, y = log2FoldChange, col = sign)) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "red", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 5e+04, y = 5, label = sprintf("Up in CompHet: %d", sum(het.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 5e+04, y = -6, label = sprintf("Down in CompHet: %d", sum(het.vs.cf$sign == "Downregulated"))) + 
  ggtitle("CompHet vs CF F508del")
dev.off()
```


## Pairwise comparisons within same k-means cluster

Then let’s run this comparison using samples from “epithelial” cluster only, which we define by deconvolution

```{r}
epi.samples <- names(kmeans.groups[kmeans.groups == 2])
```


```{r}
epi.counts <- counts[, epi.samples]
epi.coldata <- coldata[epi.samples, ]
dds <- DESeqDataSetFromMatrix(epi.counts / 2, epi.coldata, ~ genotype)
```

```{r}
dds <- DESeq(dds, fitType = "local")
```
### Control vs CF
```{r}
ctrl.vs.cf <- as.data.frame(results(dds, contrast = c("genotype", "HV", "F508del")))
ctrl.vs.cf <- ctrl.vs.cf[!is.na(ctrl.vs.cf$padj), ]
ctrl.vs.cf$sign <- ""
ctrl.vs.cf$sign[(ctrl.vs.cf$log2FoldChange < 0) & (ctrl.vs.cf$padj < 0.05)] <- "Downregulated"
ctrl.vs.cf$sign[(ctrl.vs.cf$log2FoldChange > 0) & (ctrl.vs.cf$padj < 0.05)] <- "Upregulated"
ctrl.vs.cf <- ctrl.vs.cf[order(ctrl.vs.cf$sign),]
```

```{r}
ace2_data <- as.data.frame(merge(ctrl.vs.cf[ace2_id, ], gene_table[ace2_id, ], by = "row.names", suffixes = c("", ".y")))
```

```{r}
ggplot(ctrl.vs.cf, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
    geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 5000,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  scale_fill_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 5e+04, y = 5, label = sprintf("Up in HV: %d", sum(ctrl.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 5e+04, y = -4, label = sprintf("Up in homozygous\nF508del: %d", sum(ctrl.vs.cf$sign == "Downregulated"))) + 
  ggtitle("HV vs CF F508del (Cluster A)") +
  theme(legend.position = "none")
```

Save as PDF
```{r}
pdf("04epi-samples-control-vs-f508del.pdf", width=6, height=6)
ggplot(ctrl.vs.cf, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
    geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 5000,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  scale_fill_manual(values = c("gray", "darkred", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 5e+04, y = 5, label = sprintf("Up in HV: %d", sum(ctrl.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 5e+04, y = -4, label = sprintf("Up in homozygous\nF508del: %d", sum(ctrl.vs.cf$sign == "Downregulated"))) + 
  ggtitle("HV vs CF F508del (Cluster A)") +
  theme(legend.position = "none")
dev.off()
```


```{r}
sign.only <- ctrl.vs.cf[ctrl.vs.cf$padj < 0.05, ]
sign.only$gene_name <- gene_table[rownames(sign.only), "Gene_Name"]
write.csv(sign.only[order(sign.only$log2FoldChange, decreasing = TRUE), ], "04epi-ctrl-vs-cf.csv")
```

Query GO processes for up- and down-regulated genes here
```{r}
upreg_genes <- rownames(sign.only)[sign.only$sign == "Upregulated"]
go_upreg <- go_enrichment(dds, upreg_genes, go_annotations = "org.Hs.eg.db")
```
```{r}
downreg_genes <- rownames(sign.only)[sign.only$sign == "Downregulated"]
go_downreg <- go_enrichment(dds, downreg_genes, go_annotations = "org.Hs.eg.db")
```

```{r}
go_upreg$direction <- "Upregulated"
go_downreg$direction <- "Downregulated"
go_processes <- rbind(go_upreg, go_downreg)
write.csv(go_processes, "04epi-ctrl-vs-f508del-go-processes.csv")
```


### Control vs CompHet
```{r}
ctrl.vs.het <- as.data.frame(results(dds, contrast = c("genotype", "HV", "CompHet")))
ctrl.vs.het <- ctrl.vs.het[!is.na(ctrl.vs.het$padj), ]
ctrl.vs.het$sign <- ""
ctrl.vs.het$sign[(ctrl.vs.het$log2FoldChange < 0) & (ctrl.vs.het$padj < 0.05)] <- "Downregulated"
ctrl.vs.het$sign[(ctrl.vs.het$log2FoldChange > 0) & (ctrl.vs.het$padj < 0.05)] <- "Upregulated"
ctrl.vs.het <- ctrl.vs.het[order(ctrl.vs.het$sign),]
```

```{r}
ace2_data <- as.data.frame(merge(ctrl.vs.het[ace2_id, ], gene_table[ace2_id, ], by = "row.names", suffixes = c("", ".y")))
```

```{r}
ggplot(ctrl.vs.het, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
  geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 100,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  scale_fill_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 5e+04, y = 5, label = sprintf("Up in HV: %d", sum(ctrl.vs.het$sign == "Upregulated"))) +
  annotate("text", x = 5e+04, y = -6, label = sprintf("Up in CompHet: %d", sum(ctrl.vs.het$sign == "Downregulated"))) + 
  ggtitle("HV vs CompHet (Cluster A)") +
  theme(legend.position = "none")
```

Save as PDF
```{r}
pdf("04epi-samples-control-vs-comphet.pdf", width=6, height=6)
ggplot(ctrl.vs.het, aes(x = baseMean, y = log2FoldChange)) + 
  geom_point(aes(fill = sign, color = sign), shape = 21, stroke = 0.5) +
  geom_label_repel(
    aes(label = Gene_Name, fontface = "italic"), 
    data = ace2_data,
    box.padding = 0.5,
    seed = 100,
  ) +
  geom_point(aes(fill = sign), data = ace2_data, stroke = 0.5, shape = 21) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  scale_fill_manual(values = c("gray", "darkorange3", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 5e+04, y = 5, label = sprintf("Up in HV: %d", sum(ctrl.vs.het$sign == "Upregulated"))) +
  annotate("text", x = 5e+04, y = -6, label = sprintf("Up in CompHet: %d", sum(ctrl.vs.het$sign == "Downregulated"))) + 
  ggtitle("HV vs CompHet (Cluster A)") +
  theme(legend.position = "none")
dev.off()
```


```{r}
sign.only <- ctrl.vs.het[ctrl.vs.het$padj < 0.05, ]
sign.only$gene_name <- gene_table[rownames(sign.only), "Gene_Name"]
write.csv(sign.only[order(sign.only$log2FoldChange, decreasing = TRUE), ], "04epi-ctrl-vs-het.csv")
```


### CompHet vs CF
```{r}
het.vs.cf <- as.data.frame(results(dds, contrast = c("genotype", "CompHet", "F508del")))
het.vs.cf <- het.vs.cf[!is.na(het.vs.cf$padj), ]
het.vs.cf$sign <- ""
het.vs.cf$sign[(het.vs.cf$log2FoldChange < 0) & (het.vs.cf$padj < 0.05)] <- "Downregulated"
het.vs.cf$sign[(het.vs.cf$log2FoldChange > 0) & (het.vs.cf$padj < 0.05)] <- "Upregulated"
het.vs.cf <- het.vs.cf[order(het.vs.cf$sign),]
```

```{r}
ggplot(het.vs.cf) + 
  geom_point(aes(x = baseMean, y = log2FoldChange, col = sign)) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "red", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 2e+04, y = 5, label = sprintf("Up in CompHet: %d", sum(het.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 2e+04, y = -6, label = sprintf("Down in CompHet: %d", sum(het.vs.cf$sign == "Downregulated"))) + 
  ggtitle("CompHet vs CF F508del")
```

```{r}
pdf("04epi-samples-comphet-vs-f508del.pdf", width=8, height=6)
ggplot(het.vs.cf) + 
  geom_point(aes(x = baseMean, y = log2FoldChange, col = sign)) +
  scale_x_log10() +
  scale_color_manual(values = c("gray", "red", "blue"), name = "Significance", labels = c("Not significant", "Downregulated", "Upregulated")) + 
  annotate("text", x = 2e+04, y = 5, label = sprintf("Up in CompHet: %d", sum(het.vs.cf$sign == "Upregulated"))) +
  annotate("text", x = 2e+04, y = -6, label = sprintf("Down in CompHet: %d", sum(het.vs.cf$sign == "Downregulated"))) + 
  ggtitle("CompHet vs CF F508del")
dev.off()
```


```{r}
sign.only <- het.vs.cf[het.vs.cf$padj < 0.05, ]
sign.only$gene_name <- gene_table[rownames(sign.only), "Gene_Name"]
write.csv(sign.only[order(sign.only$log2FoldChange, decreasing = TRUE), ], "04epi-het-vs-cf.csv")
```

## IFN-1/2 signatures

```{r}
ifn1.genes <- read.table("HALLMARK_INTERFERON_ALPHA_RESPONSE.txt", header = FALSE, skip = 2)
ifn1.genes <- gene_table[gene_table$Gene_Name %in% ifn1.genes$V1, "Ensembl_ID"]
ifn2.genes <- read.table("HALLMARK_INTERFERON_GAMMA_RESPONSE.txt", header = FALSE, skip = 2)
ifn2.genes <- gene_table[gene_table$Gene_Name %in% ifn2.genes$V1, "Ensembl_ID"]
```

Compute means of IFN-1/2 gene sets
```{r}
ifn1.means <- colMeans(counts_mat_orig[ifn1.genes, ])
ifn2.means <- colMeans(counts_mat_orig[ifn2.genes, ])
```

```{r}
ifn1.data <- as.data.frame(colData(dds_orig))
ifn1.data$expr <- ifn1.means[rownames(ifn1.data)]
ifn1.data$genotype <- factor(ifn1.data$genotype, levels = c("HV", "CompHet", "F508del"))
```


### IFN-I
```{r}
ggplot(ifn1.data) + 
  stat_summary(aes(x = genotype, y = expr, ymax=expr, ymin=expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(aes(x = genotype, y = expr, color = genotype), width = 0.2, height = 0) +
  xlab("") + ylab("Mean IFN-I genes expression") +
  scale_color_manual(values = c("HV" = "blue", "CompHet" = "darkorange3", "F508del" = "darkred")) +
  theme(axis.text.x = element_text(size = 14))
```

Save as PDF
```{r}
pdf("05ifn1-expr.pdf", width = 6, height = 4)
ggplot(ifn1.data) + 
  stat_summary(aes(x = genotype, y = expr, ymax=expr, ymin=expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(aes(x = genotype, y = expr, color = genotype), width = 0.2, height = 0) +
  xlab("") + ylab("Mean IFN-I genes expression") +
  scale_color_manual(values = c("HV" = "blue", "CompHet" = "darkorange3", "F508del" = "darkred")) +
  theme(axis.text.x = element_text(size = 14))
dev.off()
```

Are any significant?
```{r}
t.test(ifn1.data$expr[ifn1.data$genotype == "HV"], ifn1.data$expr[ifn1.data$genotype == "CompHet"])
```

```{r}
t.test(ifn1.data$expr[ifn1.data$genotype == "HV"], ifn1.data$expr[ifn1.data$genotype == "F508del"])
```

No

### IFN-II
```{r}
ifn2.data <- as.data.frame(colData(dds_orig))
ifn2.data$expr <- ifn2.means[rownames(ifn2.data)]
ifn2.data$genotype <- factor(ifn2.data$genotype, levels = c("HV", "CompHet", "F508del"))
```

```{r}
ggplot(ifn2.data) + 
  stat_summary(aes(x = genotype, y = expr, ymax=expr, ymin=expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(aes(x = genotype, y = expr, color = genotype), width = 0.2, height = 0) +
  xlab("") + ylab("Mean IFN-II genes expression") +
  scale_color_manual(values = c("HV" = "blue", "CompHet" = "darkorange3", "F508del" = "darkred")) +
  theme(axis.text.x = element_text(size = 14))
```

Save as PDF
```{r}
pdf("05ifn2-expr.pdf", width = 6, height = 4)
ggplot(ifn2.data) + 
  stat_summary(aes(x = genotype, y = expr, ymax=expr, ymin=expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(aes(x = genotype, y = expr, color = genotype), width = 0.2, height = 0) +
  xlab("") + ylab("Mean IFN-II genes expression") +
  scale_color_manual(values = c("HV" = "blue", "CompHet" = "darkorange3", "F508del" = "darkred")) +
  theme(axis.text.x = element_text(size = 14))
dev.off()
```

Are any significant?
```{r}
t.test(ifn2.data$expr[ifn2.data$genotype == "HV"], ifn2.data$expr[ifn2.data$genotype == "CompHet"])
```

No.

## Group IFN-1/2 expression by k-means clusters

```{r}
ifn1.data$cluster <- factor(group.names[kmeans.groups], levels = c("A", "B", "C"))
```

### IFN-I

```{r}
ggplot(ifn1.data, aes(x = cluster, y = expr, shape = cluster)) + 
  stat_summary(aes(ymax = expr, ymin = expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(width = 0.2, height = 0, size = 2) +
  xlab("") + ylab("Mean IFN-I genes expression") +
  stat_compare_means(
    comparisons = list(c("A", "B"), c("B", "C")),
    label.y = c(16000, 17000),
    method = "t.test"
  ) +
  theme(axis.text.x = element_text(size = 14))
```

Save as PDF
```{r}
pdf("06ifn1-expr-by-cluster.pdf", width = 6, height = 4)
ggplot(ifn1.data, aes(x = cluster, y = expr, shape = cluster)) + 
  stat_summary(aes(ymax = expr, ymin = expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(width = 0.2, height = 0, size = 2) +
  xlab("") + ylab("Mean IFN-I genes expression") +
  stat_compare_means(
    comparisons = list(c("A", "B"), c("B", "C")),
    label.y = c(16000, 17000),
    method = "t.test"
  ) +
  theme(axis.text.x = element_text(size = 14))
dev.off()
```


### IFN-II

```{r}
ifn2.data$cluster <- factor(group.names[kmeans.groups], levels = c("A", "B", "C"))
```

```{r}
ggplot(ifn2.data, aes(x = cluster, y = expr, shape = cluster)) + 
  stat_summary(aes(ymax = expr, ymin = expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(width = 0.2, height = 0, size = 2) +
  xlab("") + ylab("Mean IFN-II genes expression") +
  stat_compare_means(
    comparisons = list(c("A", "B"), c("B", "C")),
    label.y = c(16000, 17000),
    method = "t.test"
  ) +
  theme(axis.text.x = element_text(size = 14))
```

Save as PDF
```{r}
pdf("06ifn2-expr-by-cluster.pdf", width = 6, height = 4)
ggplot(ifn2.data, aes(x = cluster, y = expr, shape = cluster)) + 
  stat_summary(aes(ymax = expr, ymin = expr), fun = "mean", geom="crossbar", width = .5, size = .2) +
  geom_jitter(width = 0.2, height = 0, size = 2) +
  xlab("") + ylab("Mean IFN-II genes expression") +
  stat_compare_means(
    comparisons = list(c("A", "B"), c("B", "C")),
    label.y = c(16000, 17000),
    method = "t.test"
  ) +
  theme(axis.text.x = element_text(size = 14))
dev.off()
```




